Microxanox: an R package for simulating an aquatic $MICR$obial ecosystem that can occupy $OX$ic or $ANOX$ic states.

### 2023-06-29 inst/manuscript/Response.md

Editor-in-Chief

SoftwareX -->

## Response from Authors to Editor and Reviewer comments

> 0. It was difficult to follow the [Responses and Suggestions] in the new version. Those sections/paragraphs that have been rewritten, added, or modified in the paper should be included in the reply letter and explicitly highlight the parts that correspond precisely to each remark. The highlighted paper and the clean version should be shown at the end of the reply letter.

!!! Needs a response.

@@

### Reviewer #1

> 1.  Please consider reference style: Number the references (numbers in square brackets) in the list in the order in which they appear in the text.

**Response:**: Done

> 2.1 The authors in section 3.3 claim several sentences that could confuse readers. 2.1 `one discovers how the steady state of the system responds to the environmental conditions`. There are several concerns that the authors should clarify:
>
> a)  `The steady state`. Is there a unique steady state? Is there a reported contribution concerning a formal analysis of the model in the sense of stability (stable state existence and uniqueness)? Maybe, authors should refer as `a system's stable state (if it exists)`.

@@

**Response:** Thank you for this suggestions. We have changed the text to "Two numerical strategies for finding steady states and their sensitivity to parameters are implemented."

> b)  `there is no definitive best method`. Under which criteria? The authors present only two numerical evaluations. Please clarified this point.

**Response:** We have removed this claim since it is not demonstrated to be true and is unimportant.

> c)  `in order to check if results were sensitive to choice of method`. This sentence needs to be clarified. There is no comparison between numerical evaluations in any sense. The numerical simulations or simulation results are not sensitive to choice or method. Indeed, the numerical solution of the dynamical model is sensitive to a set of parameters and initial conditions. Maybe the authors mean the following: to evaluate the performance of the numerical parametric sensitivity analysis of the dynamical model concerning one parametric variation (oxygen diffusivity) under two schemes.

!!! 2.6c The comment has not been addressed.

> 2.7 It is not clear why there are two `methods` for numerical parametric sensitivity analysis. The authors should include the arguments of this.

@@

!!! 2.7 The comment has not been addressed.

> 2.8 Please consider rewrite the section 3.3 (see previous 2.1 - 2.7 points) in the sense of Parametric sensitivity analysis, Open loop dynamical behavior, Equilibrium points (see Khalil, H. K. (2002). Nonlinear Systems. Prentice Hall, New Jersey.)

**Response:** We have the opinion that the with the revisions already made and detailed above, and those below, that the terminology is now more in line with this text, and will be clearly understandable by the target audience (theoretical ecologists) even when the section is not more substatially rewritten.

!!! 2.8 The comment has not been addressed.

> 3.  Please consider the following suggestions:

!!! 3 It wasn't easy to follow the [Suggestion] in the new version.

> Line 5 `types of ecosystem`

### 2023-06-30 inst/manuscript/revision2/manuscript_microxanox_rev2.Rmd
working in revision 2

---
title: "Microxanox: an R package for simulating an aquatic $MICR$obial ecosystem that can occupy $OX$ic or $ANOX$ic states."
author:
- name: Rainer M Krug
  email: Rainer.Krug@uzh.ch;  Rainer@krugs.de
  affiliation: University of Zürich
  footnote: Corresponding Author
- name: Owen L. Petchey
  email: Owen.Petchey@ieu.uzh.ch
  affiliation: University of Zürich
address:
- code: University of Zürich
  address: Department of Evolutionary Biology and Environmental Studies, Winterthurerstrasse
    190, 8057 Zurich
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"
keywords: 
  - reproducibility
  - regime shift 
  - steady state
  - final state
  - ordinary differential equations
abstract: |
  $Microxanox$ is an R package to simulate a three functional group ecosystem (cyanobacteria, phototrophic sulfur bacteria, and sulfate-reducing bacteria) with four chemical substrates (phosphorus, oxygen, reduced sulfur, and oxidized sulfur) using a set of ordinary differential equations. Simulations can be run individually or over a parameter range.  The model can be implemented with different numbers of species per functional group. The package is constructed in such a way that the results contain the input parameter used, so that a saved results can be loaded again and the simulation be repeated. Furthermore, the package framework and code should serve as a useful starting point for making simulation models of other types of ecosystems.
journal: "SoftwareX"
date: "`r Sys.Date()`"

@@

One approach to finding how the ecosystem responds to environmental driver change is to examine the relationship between the environmental driver value and the numerical value of a system's steady state (if it exists). The first component of this approach is to run the simulation for a long time and record the final state (i.e., the state of the system at the end of a simulation). When one does this across a range of environmental conditions, one discovers how the final state of the system changes with the environmental conditions. The package contains functionality for finding final states that correspond to values of one environmental driver, namely the value of oxygen diffusivity.

When one wishes to be able to make conclusions about how the steady state of the system is affected by the environmental driver, it is very important to note that the *final state* (found by the simulation) is not guaranteed to be a *steady state*. The user must ensure that the simulation is run for sufficiently long time for any transient dynamics to disappeared, and must also check the type of long-term dynamics occurring. In the results presented here, and in the paper @Limberger2023, this was performed by visual inspection, and by checking the sensitivity of conclusions to the length of the simulation. Furthermore, the package does not include methods for a formal analysis of the stability of the system and users should take care to assess if steady states are unique.

Two numerical strategies for finding final states and their sensitivity to parameters are implemented. Two strategies are implemented in order to allow comparison of their results. The first method runs a independent simulation for each combination of initial conditions and oxygen diffusivity (we term this the *Replication method*). This is the method used in the @Bush2017 study and was used to obtain the results in figures 3 and 4 of that article.

@@

@@

The package contains a function to extract summary features of ecosystem responses to environmental change, such as the amount of hysteresis displayed by the ecosystem. Hysteresis is a key feature of ecosystem responses to environmental change, because it is related to how difficult it can be to reserve the effects of environmental change [@Scheffer2001]. The amount of hysteresis is measured as the extent of the environmental condition (here oxygen diffusitivity) for which there were two steady states. I.e., it is the extent of the environmental conditions for which historical conditions play an important role in determining the current system state.

>>>>

---
title: "Microxanox: an R package for simulating an aquatic $MICR$obial ecosystem that can occupy $OX$ic or $ANOX$ic states."
author:
- name: Rainer M Krug
  email: Rainer.Krug@uzh.ch;  Rainer@krugs.de
  affiliation: University of Zürich
  footnote: Corresponding Author
- name: Owen L. Petchey
  email: Owen.Petchey@ieu.uzh.ch
  affiliation: University of Zürich
address:
- code: University of Zürich
  address: Department of Evolutionary Biology and Environmental Studies, Winterthurerstrasse
    190, 8057 Zurich
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"
keywords: 
  - reproducibility
  - regime shift 
  - final state
  - ordinary differential equations
abstract: |
  $Microxanox$ is an R package to simulate a three functional group ecosystem (cyanobacteria, phototrophic sulfur bacteria, and sulfate-reducing bacteria) with four chemical substrates (phosphorus, oxygen, reduced sulfur, and oxidized sulfur) using a set of ordinary differential equations. Simulations can be run individually or over a parameter range.  The model can be implemented with different numbers of species per functional group. The package is constructed in such a way that the results contain the input parameter used, so that a saved results can be loaded again and the simulation be repeated. Furthermore, the package framework and code should serve as a useful starting point for making simulation models of other types of ecosystems.
journal: "SoftwareX"
date: "`r Sys.Date()`"

@@

One approach to finding how the ecosystem responds to environmental driver change is to examine the relationship between the environmental driver value and the numerical value of a system's state. The first component of this approach is to run the simulation for a long time and record the final state (i.e., the state of the system at the end of a simulation). When one does this across a range of environmental conditions, one can discovers how the final state of the system changes with the environmental conditions. The package contains functionality for finding final states that correspond to values of one environmental driver, namely the value of oxygen diffusivity.

When one wishes to be able to make conclusions about how the *steady state* of the system is affected by the environmental driver, it is very important to note that the *final state* (provided by the simulation) is not guaranteed to be a *steady state*. The software does not provide the user with a steady state. In order to somewhat safely assume that the final state is a steady state, the user must ensure that the simulation is run for sufficiently long time for any transient dynamics to disappeared, and must also check the type of long-term dynamics occurring. In the results presented here, and in the paper @Limberger2023, this was performed by visual inspection, and by checking the sensitivity of conclusions to the length of the simulation. Furthermore, the package does not include methods for a formal analysis of the stability of the system and users should take care to assess if steady states are unique.

Two numerical strategies for finding final states and their sensitivity to parameters are implemented. Two strategies are implemented in order to allow users to compare of their results. The first method runs a independent simulation for each combination of initial conditions and oxygen diffusivity (we term this the *Replication method*). This is the method used in the @Bush2017 study and was used to obtain the results in figures 3 and 4 of that article.

@@

An potentially important difference between the two methods is in the system state when a new value of oxygen diffusivity is set. In the replication method, the system state when a new value of oxygen diffusivity is set is always the same. Whereas in the temporal method, the system state when a new value of oxygen diffusivity is set is the final state of the system for the previously set value of oxygen diffusivity. Since some modellers prefer one approach and others another, we decided to implement both.

@@

The package contains a function to extract summary features of ecosystem responses to environmental change, such as the amount of hysteresis displayed by the ecosystem. Hysteresis is a key feature of ecosystem responses to environmental change, because it is related to how difficult it can be to reserve the effects of environmental change [@Scheffer2001]. The amount of hysteresis is measured as the extent of the environmental condition (here oxygen diffusitivity) for which there were two distinct final states. I.e., it is the extent of the environmental conditions for which historical conditions play an important role in determining the current system state.
